---
title: "Gapfilling of hourly PM2.5 data for all countries"
author: "Johannes Heisig"
date: "`r Sys.Date()`"
format: 
  html: default
  gfm: default
knitr:
  opts_knit:
    root.dir: "/mnt/cloud/wwu1/ec_bronze/_nogroup/ae78a1ca-a0e8-4e4e-8992-69c34947db65/UseCase_AIRCON"
---

```{r}
suppressPackageStartupMessages({
library(dplyr)
library(arrow)
library(tictoc)
})
source("R/functions.R")
station_meta = read_parquet("AQ_stations/EEA_stations_meta.parquet")
```


# Read Data

Open hourly data for all 40 countries, but filter while reading to limit memory needs (using the arrow library).

```{r}
tic()
aq = open_dataset("AQ_data/02_hourly_SSR") |> 
  filter(!is.na(PM10), !is.na(PM2.5)) |> 
  collect()
aq
toc()

nrow(aq) 
table(aq$StationType, aq$StationArea)
```


# Gapfilling Model (as proposed by Denby 2011 and Horalek et al. 2019)

Simple linear model to predict PM2.5 concentrations from PM10, population density, surface solar radiation and coordinates. Models include interactions for station type (background vs traffic) but not for station area.

```{r}
lm_pm2.5 = lm(PM2.5 ~ (PM10 + Population + SSR + Longitude + Latitude) * StationType, data = aq)

summary(lm_pm2.5)
caret::RMSE(predict(lm_pm2.5), lm_pm2.5$model$PM2.5)
caret::R2(predict(lm_pm2.5), lm_pm2.5$model$PM2.5)
```

# Prediction

Map over each country and predict PM2.5 for gaps where PM10 is available by applying the above model. The function `fill_pm2.5_gaps()` further adds a column named "pseudo" that indicates whether a value was actually measured (0) or inferred through gapfilling (1), before writing to (parquet) file.

```{r}
countries = as.character(unique(station_meta$Countrycode))
filled = purrr::map_vec(countries, fill_pm2.5_gaps, model = lm_pm2.5, 
                    gap_data = "AQ_data/02_hourly_SSR",
                    out_dir = "AQ_data/03_hourly_gapfilled")

filled |> basename()
```


